---
title: "Creating a DataSchema and config files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating a DataSchema and config files}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`cohortHarmonize` is driven entirely by three small CSV config files.
This guide walks the user through writing them. Nothing about the engine
itself changes between projects, instead the user has to alter only the
config files.

## 0. Scaffold blank templates

```{r}
library(cohortHarmonize)
new_dataschema("dataschema.csv")
new_source_map("peg_map.csv")
new_dq_metadata("dq_metadata.csv")
```

## 1. The DataSchema — deciding what the user harmonizes *to*

The DataSchema represents the shared harmonization target. One row per
harmonized variable.

Two resources about DataSchemas and their constrcution can be accessed
via:

-   <https://pubmed.ncbi.nlm.nih.gov/27272186/>

<!-- -->

-   <https://www.integrate.io/blog/complete-guide-to-database-schema-design-guide/>

| column | meaning |
|------------------------------------|------------------------------------|
| `variable` | the harmonized variable name (e.g. `age`, `exp_coffee`) |
| `domain` | a grouping used by `select_domain()` (e.g. `Demographics`) |
| `type` | `integer` / `float` (documentation + QC) |
| `description` | the **target coding** every source must land on, e.g. `1=Female, 2=Male` |

**Important:** the DataSchema should include only the constructs the
user's analysis actually needs. The target coding should be written
explicitly (what each source is re-coded to). Variables should be
grouped into domains so that downstream users can pull data from focus
areas.

## 2. A source map — connecting one source's raw columns to the schema

One source map per data source. Each row explains: *this target variable
is built from these raw column(s) using this recode rule.*

| column | meaning |
|------------------------------------|------------------------------------|
| `target_variable` | must match a `variable` in the DataSchema |
| `source_columns` | one or more raw columns, `;`-separated |
| `recode` | a rule name from `list_recodes()` (or one the user `register_recode()`) |
| `param` | optional argument to the rule (e.g. a value map `1=1;2=2`) |
| `note` | free text for the user's own documentation |

Example rows:

```         
target_variable, source_columns,                 recode,          param
sex,             gender,                          map,             1=1;2=2;3=3
age,             dob;enrollment_date,             age_from_dates,
exp_coffee,      coffee_drip;coffee_instant,      gt0,
high_bp,         high_bp;high_bp_fu,              any_of,
aspirin,         aspirin;aspirin_fu,              checkbox_any,
gds_score,       gds1;gds2;...;gds15,             (custom sum rule),
```

Validate before running:

```{r}
validate_source_map("peg_map.csv", "dataschema.csv")
```

### When a built-in rule isn't enough

The user can register a study-specific rule once, then reference it by
name:

```{r}
register_recode("race_checkbox", function(df, cols, param = NULL) {
  # cols are race___1 ... race___98; roll up to one coded race value
  # ... your logic ...
})
```

Built-ins available out of the box: `keep`, `binary`, `numeric`,
`na_codes`, `map`, `any_of`, `checkbox_any`, `gt0`, `coalesce`,
`year_of`, `age_from_dates`, `unavailable`.

## 3. dq_metadata — driving the quality report

One row per variable, describing what "valid" means so `dq_report()` can
eventually grade the harmonized output.

| column | meaning |
|------------------------------------|------------------------------------|
| `variable`, `domain` | as in the DataSchema |
| `data_type` | `integer` / `float` (integrity check) |
| `admissible_values` | allowed codes, `;`-separated (consistency: inadmissible categories) |
| `hard_limits` | `min;max` outside which a value is *inadmissible* |
| `soft_limits` | `min;max` outside which a value is *improbable* |

## 4. Running it

```{r}
ds <- "dataschema.csv"
peg <- harmonize_source(source_csv("peg.csv"), "peg_map.csv", ds, source_name = "PEG")
hbs <- harmonize_source(source_redcap(),       "hbs_map.csv", ds, source_name = "HBS")
combined <- combine_sources(peg, hbs)
dq_report(combined, "dq_metadata.csv")
```

These four steps explain the entire workflow. To add a third cohort
later, the user will have to write another source map. However, the
DataSchema, engine, and QC should stay the same.
